Analyzing Fairness: Race, Recidivism, and the COMPAS Algorithm

### 2022-04-14 Paper_Draft2/Draft2.tex
Amrita's changes

%=================================================================
% Full title of the paper (Capitalized)
\Title{Does the COMPAS Needle Always Point Towards Equity? Finding
Fairness in the COMPAS Risk Assessment Algorithm':' A Case Study}

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
01063; \\
$^{2}$ \quad ; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}}\\
$^{3}$ \quad ; \href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}}\\
$^{4}$ \quad ; \href{mailto:ekim89@smith.edu}{\nolinkurl{ekim89@smith.edu}}\\
$^{5}$ \quad ; \href{mailto:ekornberg@smith.edu}{\nolinkurl{ekornberg@smith.edu}}\\
$^{6}$ \quad ; \href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}\\
}
% Contact information of the corresponding author
\corres{Correspondence: }

@@

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{A variety of disciplines use risk assessment instruments to
help humans make data-driven decisions. Northpointe, a software company,
created an algorithmic risk assessment instrument known as the
Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS). COMPAS uses various behavioral and psychological metrics
related to recidivism to assist justice systems in assessing a
defendant's potential recidivism risk. Angwin et al.~published a
ProPublica article in which they conclude that the racial biases in the
criminal justice system are reflected in the COMPAS recidivism risk
scores. In response, Dieterich et al.~published a rebuttal on behalf of
Northpointe defending the COMPAS algorithm and refuting Angwin et al.'s
allegation of racial bias. Using a human rights framework adopted from
the organizations Women at the Table and AI Fairness 360, we use
debiasing algorithms and fairness metrics to analyze the argument
between Northpointe and ProPublica and determine whether and to what
extent there is racial bias in the COMPAS algorithm.}

@@

The Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS) algorithm was created by the private, for-profit company
Northpointe (now known by its parent company
\href{https://www.equivant.com/faq/}{Equivant}), to predict defendants'
risk of recidivism. It generates a decile score that classifies
defendants' risk of recidivism as either low, medium, or high
\citep{angwin2016machine}. Jurisdictions across the United States use
the COMPAS risk assessment instrument, including but not limited to the
New York \citep{NYDOCCS2019}, Massachusetts, Michigan, California
\citep{Jackson2020Setting}, and Wisconsin \citep{WisconsinDOC}
Departments of Corrections. Due to the proprietary nature of the COMPAS
algorithm, it is unknown how exactly these recidivism risk scores are
calculated; however, a sample COMPAS Risk Assessment Survey has been
made publicly available, revealing the algorithm's input information.
This survey has been widely critiqued for using proxy variables for race
that do not explicitly factor in a defendant's race but heavily imply
it, allowing Northpointe to claim that their algorithm is free of racial
bias. For example, the COMPAS risk assessment survey asks screeners to
speculate if a defendant might be affiliated with a gang. It also asks
if a defendant has any friends or family members who have been crime
victims \citep{Angwin2016Sample}. Although these questions do not
directly ask about race, they do not take into account the pervasive
nature of systemic racism that infiltrates every aspect of the lives of
marginalized people, thereby indirectly asking about race. Angwin et
al.~published a 2016 piece in the news publication ProPublica that
analyzes the methods and algorithms used by Northpointe in their COMPAS
risk score assessment algorithm and uncovers the racial biases in
defendants' scores \citep{angwin2016machine}. Their research finds that
``the algorithm {[}is{]} somewhat more accurate than a coin flip,'' a
worrisome level of accuracy given the potential impact its
determinations may have on real people's lives. Angwin et
al.~specifically investigate the distribution of COMPAS scores by decile
among Black and white defendants. They write: ``The analysis also
{[}shows{]} that even when controlling for prior crimes, future
recidivism, age, and gender, black defendants {[}are{]} 45 percent more
likely to be assigned higher risk scores than white defendants''
\citep{larson2016we}. After examining the fairness metric statistical
parity difference, Angwin et al.~conclude that the algorithm is racially
biased \citep{larson2016we}.

Dieterich et al., on the behalf of Northpointe, deny the allegations of
racial bias and offer their own analyses based on different fairness
metrics in rebuttal \citep{equivant_response_2016}. Angwin et
al.~maintain that there are biases in the outcome values, protected
attributes, and covariates during Dieterich et al.'s data processing
phase. ProPublica collaborators Larson et al.~account for these biases
in their analyses. In their response, Dieterich et al.~highlight that
Angwin et al.~did not account for base rates of recidivism in their
analysis, which are important to understand initial percentages without
the presence of other information.

\href{https://www.womenatthetable.net/}{Women at the Table}, the sponsor
organization for this project, is ``a growing, global gender equality \&
democracy CSO based in Geneva, Switzerland focused on advancing feminist
systems change by using the prism of technology, innovation \& AI
exercising leverage points in technology, the economy, sustainability \&
democratic governance.'' We are collaborating with the organization on
its AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked with
debiasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
corresponding data story that will be added to its library. Our project
builds on Women at the Table's various debiasing algorithms used in its
AI \& Equality Human Rights Toolbox to conduct our own analyses on the
COMPAS dataset. Based on this analysis, we employ a human rights
framework to contribute to the ProPublica and Northpointe debate and
investigate whether or to what extent there is racial bias in the COMPAS
algorithm. With a solid understanding of the two sides, we aim to
pinpoint the shortcomings of both arguments and correct them in our
analyses. We will use various debiasing techniques and fairness metrics
to evaluate the level of bias present in the COMPAS data and our
algorithm. We will summarize our results using the JupyterNotebook
framework from Women at the Table, to be used by members of the
organization to teach in a workshop setting. We hope that our findings
will highlight the importance of checking statistical analyses using
varied methods and contribute to the ongoing discussion of the effects
of machine biases in the justice system.

@@

The data we are using for this project is the COMPAS General Recidivism
Risk Scores dataset from the AI Fairness 360 (AIF360) toolkit
\citep{aif360-oct-2018}, which does the same initial preprocessing as
ProPublica. The raw data has 6,167 rows and each row represents an
arrest charge for a defendant. AIF360's COMPAS data includes the
defendant's age, race, sex, what they were charged with, and whether or
not the defendant ultimately recidivated within a two-year period after
their arrest. For the purposes of our project, which endeavors to
evaluate the differing effects of the COMPAS algorithm on white
defendants versus Black defendants, we have filtered the data to only
include individuals whose race is listed as Caucasian or
African-American. Our data therefore has 5,723 rows (Figure
\ref{fig:table snip}), with the below distributions of race, age (Figure
\ref{fig:age plot}), and two year recidivism rate (Figures
\ref{fig:recid plot} \& \ref{fig:recid race plot}).

@@

\hypertarget{methods}{%
\section{Methods}\label{methods}}

The AIF360 toolkit contains various group and individual fairness
metrics and pre-processing, in-processing, and post-processing
algorithms that we used for our experiments \citep{aif360-oct-2018}. We
researched the definitions and applications of different fairness
metrics \citep{ashokan2021fairness} to determine which metric would be
most appropriate for our project. We chose to look at four group
fairness metrics instead of individual fairness metrics. Group fairness
takes into account the attributes of a whole group as opposed to just
one individual in the group, allowing us to represent more of the
systemic issues happening. In general, group fairness metrics require
that the unprivileged group is treated similarly to the privileged
group, whereas individual fairness metrics require individuals to be
treated consistently \citep{kypraiou_what_2021}. Group and individual
metrics work in opposition of one another, meaning that when group
fairness improves, individual fairness gets worse.

\emph{Statistical Parity Difference} - This metric measures the
difference that the majority and protected classes get a particular
outcome. The ideal value of this metric is 0. Fairness for this metric
is between -0.1 and 0.1. A negative value means there is higher benefit
for the privileged group (in this case, white defendants).

\(P(\hat{Y}=1|D=Unprivileged) - P(\hat{Y}=1|D=Privileged)\)

\emph{Disparate Impact Ratio} - This metric is the ratio of how often
the favorable outcome occurs in one group versus the other. In the case
of recidivism, this is the ratio of how many white defendants are
predicted to not recidivate compared to how many black defendants are
predicted to not recidivate. A value of 1 means that the ratio is
exactly 1:1. Less than 1 means the privileged group (white defendants)
benefits, while a value greater than 1 means the unprivileged group
(Black defendants) benefits. According to AIF360, a ratio between .8 to
1.25 is considered fair:

\(\frac{P(\hat{Y}=1|D=Unprivileged)}{P(\hat{Y}=1|D=Privileged)}\)

\emph{Average Odds Difference} - This metric returns the average
difference in false positive rate and true positive rate for the
privileged and unprivileged groups. A value of 0 indicates equality of
odds, and a value below 0 implies benefit for the privileged group.
Equality of odds is achieved in the case of recidivism when the
proportion of people who were predicted to recidivate and did recidivate
is equal (true positive rate) for both Black and white defendants AND
the proportion of people who were predicted to recidivate and did not
recidivate (false positive rate) is equal for both Black and white
defendants.

\(\frac{(FPR_{D = unprivileged} - FPR_{D = privileged}) + (TPR_{D = unprivileged} - TPR_{D = privileged})}{2}\)

\emph{Equal Opportunity Difference} - This metric is computed as the
difference of true positive rates between the unprivileged and the
privileged groups. The true positive rate is the ratio of true positives
to the total number of actual positives for a given group.

The ideal value is 0. A value less than 0 implies higher benefit for the
privileged group and a value greater than 0 implies higher benefit for
the unprivileged group. Fairness for this metric is between -0.1 and
0.1. This metric is best used when it is very important to catch
positive outcomes while false positives are not exceptionally
problematic. This is not the case for the COMPAS data set, as false
positives mean extra jail time for someone who will not actually
reoffend.

\(TPR_{D = Unprivileged} - TPR_{D = Privileged}\)

For the next step of our experiment, we needed to determine where in the
data science pipeline we can mitigate the most bias, using
pre-processing, in-processing, and post-processing debiasing algorithms.
These are all based on using predictive models to figure out how we can
``fix'' the bias that is present. Pre-processing refers to assessing the
training data, and it is the most flexible method because it has not yet
trained a model that may carry assumptions about the data. It is
important to keep in mind that pre-processing prevents assumptions in
the modeling, but does not account for the bias in data collection. With
pre-processing, we use the reweighting algorithm from AIF360 which
weights the examples in each (group, label) combination differently to
ensure fairness before classification \citep{aif360-oct-2018}. We used a
logistic regression model for this algorithm, as it is the easiest to
interpret in the given context. After running the fairness metrics using
the pre-processing algorithm, we were able to compare our results to the
baseline metrics from the previous section.

Another approach, in-processing, explores how bias can be targeted while
building a model. It assesses how we build learning algorithms, which
often prioritize accuracy over fairness. The technique we use is the
Meta Classification algorithm, which accounts for the fairness metric as
part of the input and returns a classifier optimized by that particular
metric. In order to do this, we first needed to convert our data frame
into a data type called a BinaryLabelDataset. Similarly to
pre-processing, we compared the results of our in-processing methods
with both the baseline and the pre-processing to gauge which method so
far has better individual or group fairness.

Our last approach, post-processing, makes transformations on model
outputs. Similarly to the pre-processing, it is flexible because it does
not work directly with the algorithm, but instead has direct access to
all the predicted classification values. Because the algorithms are
often so complex and uninterpretable, post-processors are trained in a
black box. To get around this, we use the PostProcessingMeta class from
AIF360 which combines the training and prediction for a random estimator
and the post-processor while simultaneously splitting the dataset. Based
on these results, we looked comparatively at all the approaches to
determine which we assessed to be the most fair.

To reiterate, fairness is subjective, so the questions we ask ourselves
to create parameters for our definition of fairness include:

\hypertarget{preliminary-results}{%
\section{Preliminary Results}\label{preliminary-results}}

The statistical parity difference is -0.14. This indicates that there is
a large difference between white and Black defendants regarding whether
or not they recidivate. The algorithm unfairly benefits white defendants
over Black defendants.

Disparate impact ratio is 0.47. The ratio of white defendants predicted
to not recidivate to the Black defendants predicted to not recidivate is
0.47. A ratio between 0.8 and 1.25 is considered fair, therefore the
algorithm unfairly benefits white defendants.

Average odds difference is -0.44. The average difference in false
positive rates and true positive rates for white and Black defendants is
-0.44. Values less than zero are considered in favor of the privileged
group, so the algorithm unfairly benefits white defendants.

Equal opportunity difference is -0.41. The difference of true positive
rates between the Black and white groups is -0.41. A value less than 0
indicates a benefit to the privileged group, so the algorithm unfairly
benefits white defendants. The value is substantially less than -0.1,
which indicates that the algorithm benefits white defendants.

All four group fairness metrics determine that the COMPAS algorithm
favors white defendants over Black defendants. The statistical parity
metric indicates unfairness by a smaller margin of 0.04 than equal
opportunity difference which has a margin of error 0.4 below the
threshold. Although this is true, each algorithm uses different
penalties for unfavorable outcomes, so the higher the penalty the more
of a difference there will be between the fairness threshold and the
value.

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{It's COMPAS-licated -- Can We Find Fairness in the COMPAS Risk
Assessment Algorithm?}

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA 01063; \\
$^{2}$ \quad ; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}}\\
$^{3}$ \quad ; \href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}}\\
$^{4}$ \quad ; \href{mailto:ekim89@smith.edu}{\nolinkurl{ekim89@smith.edu}}\\
$^{5}$ \quad ; \href{mailto:ekornberg@smith.edu}{\nolinkurl{ekornberg@smith.edu}}\\
$^{6}$ \quad ; \href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}\\
}
% Contact information of the corresponding author
\corres{Correspondence: \href{mailto:leutnant@fh-muenster.de}{\nolinkurl{leutnant@fh-muenster.de}};
Tel.: +XX-000-00-0000.}

@@

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{A variety of disciplines use risk assessment instruments to help humans
make data-driven decisions. Northpointe, a software company, created a
risk assessment instrument known as the Correctional Offender Management
Profiling for Alternative Sanctions (COMPAS). COMPAS uses various
behavioral and psychological metrics related to recidivism to assist
justice systems in assessing a defendant's potential recidivism risk.
ProPublica published an article which concludes that the biases in the
criminal justice system are reflected in the COMPAS software. Using a
human rights framework adopted from the organization Women at the Table,
we use various debiasing algorithms to analyze both sides of the
argument between Northpointe and ProPublica and determine the level and
extent of racial bias in the COMPAS algorithm.}

@@

Women @ the Table, the sponsor organization for this project, is ``a
growing, global gender equality \& democracy CSO based in Geneva,
Switzerland focused on advancing feminist systems change by using the
prism of technology, innovation \& AI exercising leverage points in
technology, the economy, sustainability \& democratic governance''
\citep{noauthor_womenthetable_nodate}. We have been asked to collaborate
on their AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked
with debiasing the COMPAS algorithm
\citep{noauthor_aif360datasetscompasdataset_2018} and producing a
corresponding data story that will be added to their library.

The Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS) algorithm was created by the private for-profit company
Northpointe, also known by its parent company Equivant
\citep{equivant_faq_nodate}, to predict defendants' risk of recidivism.
It generates a score that classifies defendants' risk of recidivism as
either low, medium, or high \citep{angwin2016machine}. Jurisdictions
across the United States use the COMPAS risk assessment instrument. In
2016, ProPublica published a piece that analyzed the methods and
algorithms used by Northpointe to uncover racial biases in defendants'
scores \citep{angwin2016machine}. Their analysis looks at the
distribution of decile COMPAS scores among Black and white defendants.
ProPublica concludes, after using a statistical parity metric on false
positives or false negatives, that the algorithm is racially biased
\citep{larson2016we}. Northpointe denies the allegations of racial bias,
conducting their own analyses based on different statistical parity
metrics \citep{equivant_response_2018}. Since then, the two parties have
had several exchanges, maintaining their original arguments. Currently,
the two parties' disagreement focuses on pretrial Risk Assessment
Instruments (RAIs). ProPublica maintains that there are biases in the
outcome values, protected attributes, and covariates during
Northpointe's data processing phase. ProPublica accounts for these
biases in their analyses, and Northepointe's response, highlights how
ProPublica did not account for base rates of recidivism in their
analysis, which are important to understand initial percentages without
the presence of other information.

Our project builds on Women at the Table's various debiasing algorithms
to conduct our own analyses on the COMPAS dataset. Based on this
analysis, we employ a human rights framework to contribute to the
ongoing ProPublica and Northpointe debate and investigate whether or to
what extent there is racial bias in the COMPAS algorithm. With a solid
understanding of the two sides, we aim to pinpoint the shortcomings of
both arguments and correct them in our analyses. We will use various
machine learning algorithms including logistic regression, cross
validation, and lasso and ridge techniques to choose models. We will
summarize our results using the JupyterNotebook framework from Women at
the Table, to be used by members of the organization to teach in a
workshop setting. We hope that our findings will highlight the
importance of checking statistical analyses using varied methods and
contribute to the ongoing discussion of the effects of machine biases in
the justice system.

@@

The data we are using for this project is the COMPAS General Recidivism
Risk Scores dataset from ProPublica used in their original analysis
\citep{larson_propublicacompas-analysiscompas-scores-two-yearscsv_2022}.
The COMPAS dataset we are using is from AI Fairness 360 (AIF360) toolkit
\citep{noauthor_trusted-ai_360} which does the same initial
preprocessing as ProPublica. The raw data has 6,167 rows and each row
represents an arrest charge for a defendant. ProPublica's COMPAS data
includes the defendant's age, race, sex, what they were charged with,
and whether or not the defendant ultimately recidivated within a
two-year period after their arrest. For the purposes of our project,
which endeavors to evaluate the differing effects of the COMPAS
algorithm on white defendants and Black defendants, we have filtered the
data to only include individuals whose race is listed as Caucasian or
African-American. Our data therefore has 5,723 rows (Figure
\ref{fig:table snip}), with the below distributions of race (Figure
\ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year
recidivism rate (Figures \ref{fig:recid plot} \&
\ref{fig:recid race plot}).

@@

\hypertarget{results}{%
\section{Results}\label{results}}

### 2022-04-14 final_metrics_analysis.Rmd
Amrita's changes

from aif360.datasets import CompasDataset
from aif360.datasets import BinaryLabelDataset
from aif360.sklearn.datasets import fetch_compas
from aif360.algorithms.inprocessing import MetaFairClassifier

from aif360.sklearn.preprocessing import ReweighingMeta
from aif360.sklearn.inprocessing import AdversarialDebiasing
from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta

@@

```{python in-processing meta}
# meta = MetaFairClassifier(tau=0.8, sensitive_attr='race', type='fdr', seed=1234567)
# 
# meta.fit(train_BLD)
# y_pred_META = meta.predict(test_BLD)
# y_pred_META = y_pred_META.labels.flatten()
```

>>>>

from aif360.datasets import CompasDataset
from aif360.datasets import BinaryLabelDataset
from aif360.sklearn.datasets import fetch_compas
from aif360.algorithms.inprocessing import MetaFairClassifier, PrejudiceRemover, ARTClassifier

from aif360.sklearn.preprocessing import ReweighingMeta
#from aif360.sklearn.inprocessing import 
from aif360.sklearn.postprocessing import CalibratedEqualizedOdds, PostProcessingMeta

@@

## Prejudice Remover
```{r}
#prej = PrejudiceRemover(eta=1.0, sensitive_attr='race', class_attr='two_year_recid', seed=1234567)
#prej.fit(train_BLD)
#y_pred_PREJ = meta.predict(test_BLD)
#y_pred_PREJ = y_pred_PREJ.labels.flatten()
```

## Meta Fair Classifier 
```{python in-processing meta}
#meta = MetaFairClassifier(tau=0.8, sensitive_attr='race', type='sr', seed=1234567)
#meta.fit(train_BLD)
#y_pred_META = meta.predict(test_BLD)
#y_pred_META = y_pred_META.labels.flatten()
```
